Open-source tools in R for landscape ecology

### 2021-04-26 paper/paper.Rmd
Some layout changes

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

@@

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
Thus, the *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).

The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.

@@

Most *R* packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}

CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
*GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e. testing if functions return an expected value).

The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
For more information about *R* package development in general, see \cite{Wickham2015}.

(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language. For all major tasks, some example *R* packages are listed. For more packages related to the specific task, see Table \@ref(tab:packages).
```{r fig-workflow, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:fig1cap)"}
include_graphics("data/Figure_1.pdf")
```

@@

While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
Since publicly available data is becoming more prominent, so are *R* packages to access them.
Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{Hollister2020}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the *rgee* package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Another common spatial data processing task is required when the available data extends over a larger area than the study region.
In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
Related to that, masking certain areas of the study region using spatial filters (e.g. water bodies, urban areas) can be required.
Packages from Section \@ref(sec:spatrep) also allow for these operations.
Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the *grainchanger* package \cite{Graham2019}.
Furthermore, *landscapetools* is a collection of various utility functions for the raster data model \cite{Sciaini2018}.

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}

\begin{tabularx}{1\textwidth}{|X|}
\hline
\multicolumn{1}{|c|}{\textbf{Box 1: Where to start}} \\

@@

\hline
\end{tabularx}

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data (Fig. \@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}

The popular plotting package *ggplot2* (Fig. \@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
For more information about *ggplot2* in general, see \cite{Wickham2016}}.
The *tmap* package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) *ggplot2* package, and c) *tmap* package. All three approaches result in similar-looking outputs, but their syntax and capabilities differ. All maps show the total annual precipitation of Switzerland. Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package. The code to create the maps can be found in the Appendix.
```{r fig-map, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:capfigmap)"}
include_graphics("data/Figure_2.pdf")
```

@@

The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the *R* environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.

Most landscape metrics are represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.

@@

Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
The *spatstat* package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics. 
A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

\begin{table}

\small

@@

\end{tabularx}
\end{table}

@@

One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
Thus, the \textit{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \textit{R} environment (Fig. \@ref(fig:fig-workflow)).

The growing body of \textit{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \textit{Analysis of Spatial Data} \cite{Bivand2019a} and \textit{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \textit{R} packages related to spatial data is most likely higher.
The growing popularity of the \textit{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of \textit{R} to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \textit{R} programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.

@@

Most \textit{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two.}

CRAN (the \textit{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
\textit{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \textit{remotes} \cite{Hester2020} package.
Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e. testing if functions return an expected value).

The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
For more information about \textit{R} package development in general, see \cite{Wickham2015}.

(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the \textit{R} programming language. For all major tasks, some example \textit{R} packages are listed. For more packages related to the specific task, see Table \@ref(tab:packages).
```{r fig-workflow, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:fig1cap)"}
include_graphics("data/Figure_1.pdf")
```

@@

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by pacakges.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is \textit{raster} \cite{Hijmans2019}.
A \textit{raster} successor, \textit{terra}, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the \textit{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \textit{fasterize} \cite{Ross2020}, \textit{rasterDT} \cite{OBrien2020}, or \textit{exactextractr} \cite{Baston2020}.

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the `getData()` function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Another common spatial data processing task is required when the available data extends over a larger area than the study region.
In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
Related to that, masking certain areas of the study region using spatial filters (e.g. water bodies, urban areas) can be required.
Packages from Section \@ref(sec:spatrep) also allow for these operations.
Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \textit{grainchanger} package \cite{Graham2019}.
Furthermore, \textit{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{By being able to access these GIS software directly from within \textit{R}, these packages can allow to create easily reproducible and sharable \textit{R} scripts, even if functionality is currently only available in the corresponding GIS software.}

\begin{table}

\small

\begin{tabularx}{1\textwidth}{|X|}

\hline
\multicolumn{1}{|c|}{\textbf{Box 1: Where to start}} \\

@@

\hline
\end{tabularx}
\end{table}

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\textit{R} allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data (Fig. \@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}

The popular plotting package \textit{ggplot2} (Fig. \@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
Both, \textit{tmap} and \textit{mapview} build upon the \textit{leaflet} package and leaflet javascript library \cite{Cheng2021}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \textit{cartography} package \cite{Giraud2016}.
Further plotting methods for raster objects can be found in the \textit{rasterVis} package \cite{Lamigueiro2020}.
A slightly different approach to visualizing spatial data is adapted by the \textit{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) \textit{ggplot2} package, and c) \textit{tmap} package. All three approaches result in similar-looking outputs, but their syntax and capabilities differ. All maps show the total annual precipitation of Switzerland. Data were downloaded with the \textit{raster} package (precipitation data) and \textit{rnaturalearth} package (country borders) package. The code to create the maps can be found in the Appendix.
```{r fig-map, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:capfigmap)"}
include_graphics("data/Figure_2.pdf")
```

@@

The introduction of the \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \textit{R} environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The \textit{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
Additionally, the \textit{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.

Most landscape metrics are represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the \textit{motif} package \cite{Nowosad2021}.

@@

Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
The \textit{spatstat} package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics. 
A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.

Distance-based methods allow to detect and correct for spatial autocorrelation in data.
It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
The \textit{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
The \textit{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \textit{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
